A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-12-14 article/02_Materials_and_Methods.Rmd
better figure and table referencing

The analysis code of this paper is written in the statistics programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible: **doi**. A more in-depth description and discussion of the data preparation can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

@@

7. Manually defining a rectangular (EPSG:102013) research area with sufficient information in Central-, Northern and Northwestern Europe based on perceived spatial date distribution (Figure \ref{map}).

@@

11. Defining macro regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age (**citation**). The regions have a radius of 240km and overlap slightly (Figure **TODO**). That means some graves are counted towards not just one but two regions.

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test. The test result are listed in Table \ref{distance_and_correlation_table}.

```{r, echo = FALSE}
load("../analysis/data/output_data/distance_and_correlation_table.RData")
knitr::kable(
  distance_and_correlation_table, 
  "latex", 
  caption = "\\label{distance_and_correlation_table}Demo table", 
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace"
) %>%
  kableExtra::add_header_above(
    c(" ", "burial type" = 3, "burial construction" = 3, "variable correlation" = 2)
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position")
  ) %>%
  kableExtra::column_spec(c(2,4,5,7,9), width = "1.2cm") %>%
  kableExtra::column_spec(c(3,6,8), width = "1.8cm") %>%
  gsub("table", "table*", .)
```

Defining a measure of spatial distance for the above established main units of observation, the macro regions, is only possible in a highly simplified way. A continuous distance measure can't be employed, but it's possible to distinguish ordinal distance classes. For the eight regions four different categories of spatial distance were defined. The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the difference in the data level of measurements (Figure **TODO**).

To get a better understanding how the correlation coefficient should develop if there were strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g$ groups with a group population of $N_g$ each. At the beginning of the simulation there are $k$ ideas randomly, but equally distributed in this population. With every timestep -- $t$ defines the total number of time steps -- all individuals learn from an other randomly chosen individual. The probability that this other individual is from an other group equals $m_i$, for the opposite therefore $1-m_i$. The partner group for each individual learning instance is selected randomly following a group interaction probability defined in the matrix $I$. The distribution of the $k$ variants can be observed for every timestep and documented in a time series like the one for real world data (Figure **TODO**).

>>>>

The analysis code of this paper is written in the statistics programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible. A more in-depth description and discussion of the data preparation can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

@@

7. Manually defining a rectangular (EPSG:102013) research area with sufficient information in Central-, Northern and Northwestern Europe based on perceived spatial date distribution (Figure \ref{map} A).

@@

11. Defining macro regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age (**citation**). The regions have a radius of 240km and overlap slightly (Figure \ref{map} B). That means some graves are counted towards not just one but two regions.

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test.

Defining a measure of spatial distance for the above established main units of observation, the macro regions, is only possible in a highly simplified way. A continuous distance measure can't be employed, but it's possible to distinguish ordinal distance classes. For the eight regions four different categories of spatial distance were defined (Figure \ref{map} C). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the difference in the data level of measurements. The test result are listed in Table \ref{tab:distanceandcorrelationtable*}.

```{r distanceandcorrelationtable, echo=FALSE, tidy=FALSE}
load("../analysis/data/output_data/distance_and_correlation_table.RData")
knitr::kable(
  distance_and_correlation_table, 
  "latex", 
  caption = "Mean cultural distance, correlation and mantel test results", 
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace"
) %>%
  kableExtra::add_header_above(
    c(" ", "burial type" = 3, "burial construction" = 3, "variable correlation" = 2)
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position")
  ) %>%
  kableExtra::column_spec(c(2,4,5,7,9), width = "1.2cm") %>%
  kableExtra::column_spec(c(3,6,8), width = "1.8cm") %>%
  gsub("table", "table*", .)
```

To get a better understanding how the correlation coefficient should develop if there were strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g$ groups with a group population of $N_g$ each. At the beginning of the simulation there are $k$ ideas randomly, but equally distributed in this population. With every timestep -- $t$ defines the total number of time steps -- all individuals learn from an other randomly chosen individual. The probability that this other individual is from an other group equals $m_i$, for the opposite therefore $1-m_i$. The partner group for each individual learning instance is selected randomly following a group interaction probability defined in the matrix $I$. The distribution of the $k$ variants can be observed for every timestep and documented in a time series like the one for real world data.
